---
title: "Getting started with coreval"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with coreval}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

If you build SDTM datasets, you know the loop: write code, export, upload to a
validation tool, wait, read the report, work out which line of code caused each
finding, fix, repeat.

Hardly any of that time is spent *fixing* things. It goes on **finding out
what's broken**.

coreval does the finding part on your machine, in seconds. You still run your
qualified tool before you submit — you just arrive with a lot less for it to
find.

> coreval is a personal open-source project. It's not a CDISC product, isn't
> affiliated with or endorsed by CDISC, and isn't qualified or validated
> software. Treat every result as a hint, not a verdict. Your qualified tool
> and your own review are still what decide whether data is good to go.

```{r setup}
library(coreval)
```

## Start with one dataset

This is the one you'll use most while writing code. You have a data frame;
check it.

Note row 2: `2024-02-30`. February never has 30 days.

```{r one-dataset}
ae <- data.frame(
  STUDYID = "DEMO", DOMAIN = "AE", USUBJID = c("S1", "S1", "S2"),
  AESEQ = c(1, 2, 1), AETERM = c("Headache", "Nausea", "Rash"),
  AESTDTC = c("2024-01-10", "2024-02-30", "2024-01-12"),
  AEENDTC = c("2024-01-12", "2024-02-01", "")
)

result <- check_dataset(ae)
result
```

Each problem is described in words, with the rows and values that caused it, and
the rule number at the end in case you want to look it up. The worst offenders
come first, so the top of the report is where to start.

When you want the rows themselves — to filter or count — they are in
`result$findings`, with the same description in an `issue` column:

```{r one-dataset-date}
result$findings[result$findings$Value == "2024-02-30", ]
```

You can pass a file instead of a data frame — `.xpt`, `.sas7bdat` or `.csv`:

```{r one-dataset-file, eval = FALSE}
result <- check_dataset("ae.xpt")
```

coreval works out the domain from your `DOMAIN` column, or from the file name if
there isn't one. A split dataset in `ae1.xpt` still gets checked as `AE`. If it
guesses wrong, just say so: `check_dataset(ae, domain = "AE")`.

### What it couldn't check, and why that matters

Lots of CDISC rules compare **one dataset against another** — an adverse event
date against the subject's reference dates in `DM`, a visit against the trial
design. Give coreval a single dataset and those questions simply can't be
answered.

coreval won't guess. It skips them and tells you what it wanted:

```{r one-dataset-skips}
cross <- result$skipped[grepl("was not supplied", result$skipped$reason), ]
nrow(cross)
head(unique(cross$reason), 3)
```

If it ran those anyway, it would be comparing your data against columns that
aren't there, and reporting problems that don't exist. Saying nothing is better
than making something up.

Most rules do still run — across AE, DM, LB and VS, 76–84% of the applicable
ones work on a single dataset. But the ones that can't are the cross-dataset
checks, and those are often the ones you care about.

**So a short findings list here doesn't mean your data is clean.** It's a quick
first pass, not a verdict.

## Then check the whole study

Once the datasets exist as files, point coreval at the folder. Here's a small
one, built on the fly so this vignette runs without any data of your own:

```{r build-study}
dir <- tempfile("coreval_demo_")
dir.create(dir)

dm <- data.frame(
  STUDYID = "DEMO", DOMAIN = "DM", USUBJID = c("S1", "S2", "S3"),
  RFSTDTC = c("2024-01-05", "2024-01-06", ""),
  AGE = c(34, 61, 47), AGEU = c("YEARS", "YEARS", ""),
  SEX = c("M", "F", "F")
)

haven::write_xpt(dm, file.path(dir, "dm.xpt"))
haven::write_xpt(ae, file.path(dir, "ae.xpt"))
```

`read_study()` wants a **folder**, not a file, and reads everything in it:

```{r read}
study <- read_study(dir)
names(study$datasets)
```

That's the point of reading it all at once — now the cross-dataset rules have
both halves to work with. If there's a Define-XML in the folder, coreval finds
it and uses it (that needs the `xml2` package).

Then one call does the rest:

```{r check}
study_result <- check_study(study)
```

## Reading the results

You get two tables back, and you want to look at both.

### What's wrong

```{r findings-head}
head(study_result$findings)
```

One row per problem, pointing straight at it:

| Column | What it tells you |
|---|---|
| `rule_id` | which CDISC rule fired |
| `Dataset` | which dataset, or `STUDY` for whole-study checks |
| `Record` | row number, counting from 1 |
| `Variable` | the variable being complained about |
| `Value` | what was actually in there |

Two things that surprise people at first. Findings come in rule order, not
worst-first. And `Not in dataset` under `Value` means the rule wanted a variable
you don't have — which is usually exactly the finding.

It's a plain data frame, so slice it however you like:

```{r findings-filter}
f <- study_result$findings
head(f[f$Dataset == "DM", ])
sort(table(f$rule_id), decreasing = TRUE)[1:3]
```

### What couldn't be checked

```{r skipped}
head(study_result$skipped)
```

**This is the table people skip, and it's the one that bites.** An empty
findings table means one of two things: your data is clean, or a lot of rules
never ran. Those look identical if you only read the findings. coreval always
shows you both, with a reason for every rule it couldn't run.

### Saving it, and tracking what you didn't fix

```{r export, eval = FALSE}
write_findings(study_result, "issues.xlsx")  # one workbook, two sheets
write_findings(study_result, "issues.csv")   # issues.csv + issues_skipped.csv
```

Both tables get written every time, for the reason just above. Excel output
needs the `writexl` package.

The saved file has three empty columns — `Status`, `Owner` and `Notes` — for you
to fill in once it is open. Not every finding is something you will fix: some are
expected, some belong to someone else, some are waiting on a data query. Those
decisions are worth recording next to the finding rather than in a separate
document. Pass `tracking = FALSE` if you would rather not have them.

## The rule set

```{r rules}
rules_version()

rules <- list_rules()
nrow(rules)
table(rules$source)
```

`rules_version()` is the exact CDISC commit the bundled rules came from — worth
writing down next to your results.

Not every rule carries the same weight. `source` separates fully-vetted
published rules from deprecated and draft ones; `NOTICE.md` explains each.

```{r rules-for-domain}
ae_rules <- rules_for_domain("AE")
nrow(ae_rules)
```

## Two last things

**Nothing leaves your machine.** No internet, no API key, no account. The rules
and the standards metadata are bundled inside the package.

**Think of the accuracy number as a floor.** coreval agrees with CDISC's own
reference results on about 95% of the published rules that ship reference data.
But that reference data is mostly small, single-file datasets, so it doesn't
exercise much of what real submissions do. Agreement is a lower bound on
correctness, not a quality score — which is why the advice stays the same: run
your qualified tool before you submit.

```{r cleanup, include = FALSE}
unlink(dir, recursive = TRUE)
```
