---
title: "Getting started with coreval"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with coreval}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this is for

If you prepare clinical trial data for submission, you already have a way to
check it for conformance. It probably involves exporting transport files,
uploading them somewhere, waiting, reading a report, and copying findings back
to the code that produced them. Then you fix one thing and do it again.

`coreval` collapses the *discovery* half of that loop into a function call that
runs where your code already lives. It is a fast, local, **unqualified**
pre-check: something to run while you are still writing the code, so that the
formal validation run has less to find.

It does not replace a qualified validation system, and it isn't trying to.

> **coreval is an independent, personal open-source project.** It is not a
> CDISC product, is not affiliated with or endorsed by CDISC, and is not
> qualified or validated software. It evaluates a copy of the publicly
> published CDISC Open Rules so you can catch problems early and cheaply. Every
> result it gives you is an *indication*, not a certification — your qualified
> tool and your own review remain the authority on whether data is submittable.

```{r setup}
library(coreval)
```

## A worked example

Everything below runs against a small study built on the fly, so you can
execute this vignette without any data of your own.

```{r build-study}
dir <- tempfile("coreval_demo_")
dir.create(dir)

dm <- data.frame(
  STUDYID = "DEMO", DOMAIN = "DM", USUBJID = c("S1", "S2", "S3"),
  RFSTDTC = c("2024-01-05", "2024-01-06", ""),
  AGE = c(34, 61, 47), AGEU = c("YEARS", "YEARS", ""),
  SEX = c("M", "F", "F")
)
ae <- data.frame(
  STUDYID = "DEMO", DOMAIN = "AE", USUBJID = c("S1", "S1", "S2"),
  AESEQ = c(1, 2, 1), AETERM = c("Headache", "Nausea", "Rash"),
  AESTDTC = c("2024-01-10", "2024-02-30", "2024-01-12"),
  AEENDTC = c("2024-01-12", "2024-02-01", "")
)

haven::write_xpt(dm, file.path(dir, "dm.xpt"))
haven::write_xpt(ae, file.path(dir, "ae.xpt"))
```

Note `AESTDTC` on the second adverse event: `2024-02-30` is not a real date.

### 1. Read the study

`read_study()` takes a **folder**, not a single file, and reads every dataset in
it.

```{r read}
study <- read_study(dir)
names(study$datasets)
```

Reading the whole folder at once matters: many rules compare *across* datasets,
so a rule checking an adverse event date against the subject's reference dates
in `DM` can only work if both are present.

If a Define-XML file sits in the folder it is picked up automatically, and the
rules that compare your data against what Define declares become available.
That needs the `xml2` package.

### 2. Run the checks

```{r check}
result <- check_study(study)
```

One call runs every rule that applies to each dataset, plus the cross-dataset
and whole-study rules.

### 3. Read the findings

There is one row per *(dataset, record, variable)*:

| Column | Meaning |
|---|---|
| `rule_id` | which CDISC rule fired |
| `Dataset` | which dataset, or `STUDY` for whole-study checks |
| `Record` | row number in that dataset, counting from 1 |
| `Variable` | the variable the rule reports on |
| `Value` | what was actually there |

The impossible date we planted is in there. `Record` counts from 1 in the
dataset as read, so it points straight at the offending row:

```{r findings-date}
result$findings[result$findings$Value == "2024-02-30", ]
```

`CORE-000547` is the ISO 8601 rule: date variables must hold real dates, and
there is no 30 February.

Findings arrive in rule order, not severity order, and a small study still
trips a fair number of rules — mostly about variables that a real submission
would have and this three-row demo does not:

```{r findings-head}
head(result$findings)

nrow(result$findings)
length(unique(result$findings$rule_id))
```

`Not in dataset` in `Value` means the rule reports on a variable this dataset
does not have, which is itself usually the finding.

Because it is a plain data frame, narrowing it needs nothing special — one
dataset, one rule, or only the record-level findings:

```{r findings-filter}
f <- result$findings
head(f[f$Dataset == "DM", ])
head(f[!is.na(f$Record), ])
```

### 4. Always look at what was skipped

```{r skipped}
head(result$skipped)
```

This is the step people forget. A short findings table can mean your data is
clean, or it can mean a lot of rules never ran — and those look identical if you
only read `findings`. A rule `coreval` cannot evaluate is always reported here
with a reason, never counted as a pass.

### 5. Export it

```{r export, eval = FALSE}
write_findings(result, "issues.xlsx")  # one workbook, two sheets
write_findings(result, "issues.csv")   # issues.csv + issues_skipped.csv
```

Both tables are always written, for the reason above. Excel output needs the
`writexl` package.

## Checking one dataset while you write it

Everything above assumes a study folder. That is the right shape once a study
exists, but it is the wrong shape while you are still writing the code that
produces a single domain. `check_dataset()` takes the data frame you already
have:

```{r one-dataset}
ae_only <- data.frame(
  STUDYID = "DEMO", DOMAIN = "AE", USUBJID = c("S1", "S1"),
  AESEQ = c(1, 2), AETERM = c("Headache", "Nausea"),
  AESTDTC = c("2024-01-10", "2024-02-30")
)

quick <- check_dataset(ae_only)
quick$findings[quick$findings$Value == "2024-02-30", ]
```

It accepts a path to a single `.xpt`, `.sas7bdat` or `.csv` file too. The domain
is read from the `DOMAIN` column, falling back to the file name, so `ae1.xpt`
from a split dataset is still checked as `AE`; pass `domain =` to override.

The important part is what it refuses to do. A rule that needs another dataset
cannot be answered from one, so it is **not run** — it is reported with the
domain it would have needed:

```{r one-dataset-skips}
cross <- quick$skipped[grepl("was not supplied", quick$skipped$reason), ]
nrow(cross)
head(unique(cross$reason), 3)
```

Evaluating those anyway would compare against columns that are absent and
report findings built out of nothing, which is worse than saying nothing.

Most rules still run — across AE, DM, LB and VS, 76–84% of the applicable rules
execute on the dataset alone. But the ones that cannot are precisely the
cross-dataset checks, so treat this as a fast first pass, not a verdict, and run
`check_study()` on the whole folder before concluding anything.

## Working with the rule set

```{r rules}
rules_version()

rules <- list_rules()
nrow(rules)
table(rules$source)
```

`rules_version()` is the exact upstream commit the bundled rules came from —
worth recording alongside results.

Not every rule carries equal weight. `source` distinguishes fully-vetted
published rules from deprecated and draft ones; see `NOTICE.md` for what each
means.

```{r rules-for-domain}
ae_rules <- rules_for_domain("AE")
nrow(ae_rules)
```

## Some things worth knowing

**Findings are a plain data frame.** Filter, count and join them however you
normally would.

```{r summarise}
sort(table(result$findings$rule_id), decreasing = TRUE)[1:3]
```

**Nothing leaves your machine.** No internet, no API key, no account. The rules
and the standards metadata are bundled inside the package.

**It is a floor, not a ceiling.** `coreval` agrees with CDISC's own reference
results on about 95% of the published rules that ship reference data — but that
data is mostly simple, single-file datasets, so agreement is a lower bound on
correctness rather than a quality score. Run your qualified tool before you
submit.

```{r cleanup, include = FALSE}
unlink(dir, recursive = TRUE)
```
